---
title: "Assignment 4: Simpson's Index Function"
author: "Emma"
format: html
editor: visual
---

```{=html}
<style>
body {
  background-color: #f4f8f4;
}

h1 {
  color: #2f5d50;
  font-weight: bold;
}

h2 {
  background-color: #7aa095;
  color: white;
  padding: 6px;
  border-radius: 6px;
}

h3 {
  color: #2f5d50;
}

code {
  color: #2f5d50;
}
</style>
```

## Intro

In this assignment, I created a function in R to calculate Simpson’s Index of biodiversity. Simpson’s Index is used to measure biodiversity by considering both the number of species present and the abundance of each species within a sample. The function takes a vector of species names as input, counts the number of individuals in each species, and applies Simpson’s Index formula to calculate a biodiversity value between 0 and 1. I then tested the function using multiple example datasets to compare how different species distributions affect biodiversity.

## Notes from Slides

-   Functions in R should include comments both at the top of the function and within the function body to explain what the code is doing.

-   Function documentation can later be used to automatically generate documentation when building R packages.

-   Functions are often saved as separate `.R` files so they can be reused in multiple scripts or projects.

-   Function files are commonly stored inside an `R` directory within a project.

-   The `source()` function can be used to load external function files into a Quarto document or R script.

-   The lecture example used `power_gen_orig.R` as a template for proper function formatting and documentation.

-   Using functions helps organize code, improve readability, and make analyses easier to repeat and modify.

## Simpson Index

```{r}
#' Simpson's Index Function
#'
#' This function calculates Simpson's Index of biodiversity.
#'
#' @param species_vector vector containing species names for each individual
#'
#' @return Simpson's Index value between 0 and 1
#'
#' Uses the formula:
#' D = sum((ni / N)^2)
#'
#' ni = number of individuals in each species
#' N = total number of individuals

simpson_index = function(species_vector) {
  
  # convert species vector into a factor
  species_factor = factor(species_vector)
  
  # count number of individuals in each species
  species_counts = table(species_factor)
  
  # calculate total number of individuals
  N = sum(species_counts)
  
  # calculate Simpson's Index
  D = sum((species_counts / N)^2)
  
  # return biodiversity index
  return(D)
}
```

# Assignment 4

## Function Contract

This function calculates Simpsons Index of biodiversity.

The input for the function is a vector containing species names for each individual in a sample.

The output of the function is a numeric value between 0 and 1 that represents Simpsons Index.

Simpsons Index:

\[ D = \sum ((n_i/N)\^2) \]

where:

-   (n_i) = number of individuals in each species
-   \(N\) = total number of individuals

The function first converts the species vector into a factor, then counts the number of individuals in each species using `table()`. After that, it calculates the total number of individuals and computes Simpsons Index.

## Function

```{r}

## Test 1

species_example <- c(
  "oak", "oak", "oak",
  "pine", "pine",
  "maple",
  "redwood"
)

simpson_index(species_example)


```

```{r}


## Test 2

species_example2 <- c(
  "species_a", "species_a", "species_a",
  "species_b", "species_b", "species_b",
  "species_c", "species_c", "species_c"
)

simpson_index(species_example2)

```

```{r}

## Test 3

species_example3 <- c(
  "fish",
  "crab",
  "shark",
  "eel",
  "octopus"
)

simpson_index(species_example3)
```

In this assignment, I created a function in R to calculate Simpson’s Index of biodiversity. The function takes a vector of species names, counts the number of individuals in each species, and calculates the biodiversity index using Simpson’s formula. After testing the function with multiple example datasets, I found that samples with a more even distribution of species had lower Simpson’s Index values, while samples dominated by fewer species had higher values.
